---
title: "Data I/O Lab, Part 1"
author: "BIOST 544"
output: html_document
---

```{r, include = TRUE}
# install.packages("readxl")
library(readr) 
library(readxl)
```

This lab uses different versions of the `iris` dataset. For each question, click the `iris` file in the Files pane to open it in an RStudio window.

You can see more about this dataset by running `?iris` in the console.

1.  Read in the `iris_q1.csv` dataset into the `iris_q1` R object. What delimiter separates columns?

```{r}
getwd()
# read iris data with delimiter ","
iris_q1 <- read_csv("../data/iris/iris_q1.csv")
str(iris_q1)
```

2.  Read in the `iris_q2.tsv` dataset into the `iris_q2` R object. What delimiter separates columns?

```{r}
# read iris data with delimiter "\t"
iris_q2 <- read_tsv("../data/iris/iris_q2.tsv")
str(iris_q2)
```

3.  Read in the `iris_q3.txt` dataset into the `iris_q3` R object. What delimiter separates columns?

```{r}
# read iris data with delimiter " "
iris_q3 <- read_table("../data/iris/iris_q3.txt")
str(iris_q3)
```

4.  Read in the `iris_q4.csv` dataset into the `iris_q4` R object. What delimiter separates columns? [hint: file extension is ambiguous]

```{r}
# read iris data with delimiter ":"
iris_q4 <- read_delim("../data/iris/iris_q4.csv", delim = ":")
str(iris_q4)
```

5.  Read in the `iris_q5.csv` dataset into the `iris_q5` R object. Some entries read "-999" but are supposed to be missing (the value `NA` in the data).

```{r}
# read iris data with delimiter ":" and missing data encoded as "-999"
iris_q5 <- read_csv("../data/iris/iris_q5.csv", na = c("", "NA", "-999"))
str(iris_q5)
```

6.  Read in sheet 2 of the `iris_q6.xlsx` dataset into the `iris_q6` R object. How many rows are in the dataset?

```{r}
# read iris data formatted as a Microsoft Excel file
iris_q6 <- read_xlsx("../data/iris/iris_q6.xlsx", sheet = 2)
str(iris_q6)
```

```{r}
read_delim("../data/iris/iris_q1.csv")
read_delim("../data/iris/iris_q2.tsv")
read_delim("../data/iris/iris_q3.txt")
read_delim("../data/iris/iris_q4.csv")
read_delim("../data/iris/iris_q5.csv", na = c("", "NA", "-999"))
read_delim("../data/iris/iris_q6.xlsx")
```

