---
title: "Data I/O Lab, Part 2"
author: "BIOST 544"
output: html_document
---

In this lab you can use the interactive console to explore but please record your commands here.  Remember anything you type here can be "sent" to the console with Cmd-Enter (OS-X) or Ctrl-Enter (Windows/Linux) (But only in side the ```{r}``` areas).

```{r, message = FALSE}
library(dplyr)
library(tidyverse)
library(naniar)
library(visdat)
#install.packages("dplyr")
```


1. Read in the the Charm City Circulator Dataset using `read_csv`, call it `circ`
```{r}
circ <- read_csv("../data/Charm_City_Circulator_Ridership.csv")
```

2. Print the number of rows of `circ`
```{r}
nrow(circ)
```

3. Visualize the missing data patterns of `circ` with `vis_miss`:
```{r}
# count number of missing observations in each variable
colSums(is.na(circ))
# count percent of missing missing observations 
round(colSums(is.na(circ)) / nrow(circ), 2)

# visualize percent of missing observations
visdat::vis_miss(circ, sort_miss = TRUE)
```

4. Determine the percent missing in each column using `miss_var_summary`

```{r}
naniar::miss_var_summary(circ, order = TRUE)

```

5. Determine the number missing in each column using `gg_miss_var`:
```{r}
naniar::gg_miss_var(circ)

```

6. Write out `circ` to a file called "Circulator.csv" using `write_csv`:
```{r}

```

7. Write out `circ` to a file called "Circulator.rds" using `write_rds`:
```{r}

```

