---
title: "BIOST 544: Homework 1"
subtitle: "Department of Biostatistics @ University of Washington"
author: 
- Alejandro Hernandez
date: "October 5, 2024"
output: pdf_document
---

```{r setup, include=F}
# setup options
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
options(knitr.kable.NA = '-')
labs = knitr::all_labels()
labs = labs[!labs %in% c("setup", "llm_appendix", "allcode")]
```

```{r load-libraries-data, include=F}
# clear environment
rm(list=ls())

# load relevant packages
# library(dplyr)      # data frame manipulation
# library(knitr)      # table formatting
# library(gtsummary)  # "table 1" summary
# library(tidyverse)

# load data
nsclc <- read.table("../data/nsclc-modified.txt")
```

## Background

There is a belief that the effectiveness of the anti-angiogenesis agent TFD725 (evaluated in the **nsclc** dataset) may be different for older vs younger patients. We will use the nsclc dataset to attempt to evaluate this.

## Analysis

(1) As a first pass, we will consider a few subgroups of patients: Those 50 and older (50+), 55+, 60+, 65+, and 70+. Please estimate/evaluate the probability a patient on TFD725+docetaxel will survive past 400 days in each of those subgroups. Please also give an interval estimate for each of those probabilities.

```{r question-1}
# create a factor variable from patient age
nsclc$age2 <- cut(nsclc$age, 
                  c(seq(50,70,5), max(nsclc$age)), include.lowest = TRUE,
                  labels = c("50+","55+","60+","65+","70+"))
# remove the few observations whose age was below 50 (their age2 is NA)
nsclc <- subset(nsclc, !is.na(age2))
table(nsclc$age2) # number of patients within each group
# there are too few patients aged 70+ for it to be a suitable category

# perform logistic regression to estimate the age-specific treatment effects
mod.lr <- glm(survival.past.400 ~ tx*age2, data=nsclc, family=binomial())
coef(mod.lr)
# confint(mod.lr)
```

2) Now, in each of those subgroups evaluate whether TFD725+docetaxel is more effective than docetaxel alone (and the magnitude of any potential treatment effect). In addition, evaluate if the treatment effect appears to substantively and/or systematically differ across age (or if the data doesn’t give a clear answer to this).

```{r question-2}

```


**End of report. Code appendix begins on the next page.**

\pagebreak

## Code Appendix

```{r allcode, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```

**End of document.**